Gravitational Lens Modelling in a Citizen Science Context

### 2014-09-22 text/ms.tex
edits

We develop a method to enable collaborative modelling of gravitational
lenses and lens candidates by non-professional lens enthusiasts.  It
uses an existing free-form modelling program (GLASS), but
enables the input to this code to be provided in a novel way,
via a user-provided diagram that is essentially a sketch of an
arrival-time surface.

An implementation of this method, \spl, is tested in a modelling
challenge using 29 simulated lenses drawn from the set created
for the \sw citizen-science lens search.  We find that
volunteers from this online community asserted the image
parities and time ordering consistently in some lenses, but made
errors in other lenses depending on the image morphology. While
errors in image parity and time ordering lead to  large errors in the
mass distribution, the enclosed mass was found to be more
robust: the model-derived Einstein radii found by the volunteers were
consistent with those produced by one of the professional team,
suggesting that given the appropriate tools, gravitational lens
modelling is a data analysis activity that can be crowd-sourced to
good effect. Ideas for improvement are discussed; these include
(a)~overcoming the tendency of the models to be shallower than the
correct answer in test cases, leading to systematic over-estimation of
the Einstein radius by 10\% at present, and (b)~detailed modelling of
arcs.

@@

The first two discoveries of gravitational lenses
\citep{1979Natur.279..381W,1980Natur.285..641W}, where a galaxy images
a single background quasar into two (four), brought in their wake the
first work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y}.

%CM:
%The first work on lens modelling (Young et al. 1981a,b) was developed 
%after the discovery of the first two gravitational lenses (Walsh et al. 
% 1979; Weymann et al. 1980), where a massive galaxy causes a background 
% quasar to be presented as two or four images.

For the second lens
(PG1115+080), mass models scored an early success with the prediction
that one of the lensed images seen would split further into two at
higher resolution.  That galaxies must sometimes cause multiple images
had long been expected \citep{1937ApJ....86..217Z}, and it had even
been argued that the phenomenon could help measure cosmological
parameters \citep{1964MNRAS.128..307R,1966MNRAS.132..101R}, but
apparently nobody was expecting that lenses would need detailed
modelling.  The first observations, however, immediately stimulated
models.  The reason for that lies in the image separation. Recall that
image separations are of order of the angular Einstein radius
\begin{equation}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2},
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  A lensing
galaxy with $M\sim10^{11}M_\odot$ at $10^9\rm\,pc$ would cause image
separations of $\sim1''$, which is comparable to the size of the
galaxy; typically the lensed images are seen through the galaxy halo.
Hence, the lensed images depend on the detailed mass distribution of
the lensing galaxy.  Galaxy lenses demand modelling.

Since the early discoveries, more than 400 secure lenses are now
known. Modelling of the mass distribution is part of any research
using lenses, but so far no modelling study has spanned all known
lenses.  The largest single one \citep{2009ApJ...703L..51K} models 58
lenses to infer the distribution of dark matter around galaxies.  In
other work, \cite{2011ApJ...740...97L} combined lens models of 21
galaxies with models of their stellar populations, to find the
relation between stars and dark matter, and \cite{2014MNRAS.437..600S}
modelled 18 time-delay lenses together to infer cosmological
parameters.

@@

Consider a lens at some redshift $z_L$ and let $(x,y)$ be planar
coordinates at the lens, transverse to the line of sight.  Let
$\Sigma(x,y)$ be the mass per unit area, i.e., density projected along
the line of sight.  This surface density is often given in a
dimensionless form
\begin{equation} \label{eq:kappa}
\kappa(x,y) \propto \Sigma(x,y)
\end{equation}
called the convergence.  Let there be light, in the form of a more
distant source, at redshift $z_S$, behind point $(x_s,y_s)$ on the
lens.

@@

After examining this feedback, the user can choose to save the
model to the \spl archive, at which point it is assigned an unique
URL.  He can also modify his input and try again, or discard the
attempt altogether.  After archiving, there can be discussion among
modellers, through the \sw forum or by any other means, and revision
of the model.  This is achieved simply by sharing the model's
URL; following its hyperlink takes one to the \spl app, pre-loaded
with the correct data image and input spaghetti. Any archived model
can be revised by any user: they can modify the spaghetti
configuration slightly or drastically, or change options like the size
of the mass tiles. Particularly interesting lens candidates lead to
trees of models in this way.  Discussion among modellers tends to
prune a model tree, focusing attention on the most interesting
models.
% PJM: sadly this Letter is no longer available, Julianne took it
% down. We can post the talk URL instead.
% \footnote{See ``Collaborative gravitational lens
% modelling\dots'' in {\tt http://letters.zooniverse.org} for an
% example.}
\footnote{See the ``collaborative modelling'' thread at
\url{ http://talk.spacewarps.org} for an example.}

>>>>

We develop a method to enable collaborative modelling of
gravitational lenses and lens candidates, that could be used by
non-professional lens enthusiasts.  It uses an existing free-form
modelling program (GLASS), but enables the input to this code to be
provided in a novel way, via a user-provided diagram that is
essentially a sketch of an arrival-time surface.

An implementation of this method, \spl, is tested in a modelling
challenge that used 29 simulated lenses that had been drawn from the
set created for the \sw citizen-science lens search.  We find that
volunteers from this online community asserted the image parities and
time ordering consistently in some lenses, but made errors in other
lenses depending on the image morphology. While errors in image parity
and time ordering lead to large errors in the mass distribution, the
enclosed mass was found to be more robust: the model-derived Einstein
radii found by the volunteers were consistent with those produced by
one of the professional team, suggesting that given the appropriate
tools, gravitational lens modelling is a data analysis activity that
can be crowd-sourced to good effect. Ideas for improvement are
discussed; these include (a)~overcoming the tendency of the models to
be shallower than the correct answer in test cases, leading to
systematic over-estimation of the Einstein radius by 10\% at present,
and (b)~detailed modelling of arcs.

@@

The first two discoveries of gravitational lenses
, where a galaxy images
a single background quasar into two (four), brought in their wake the
first work on lens modelling
.

%CM:
The first work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y} was developed after
the discovery of the first two gravitational lenses
\citep{1979Natur.279..381W,1980Natur.285..641W}, where a massive
galaxy causes a background
quasar to be presented as two or four images.

For the second lens
(PG1115+080), mass models scored an early success with the prediction
that one of the lensed images seen would split further into two at
higher resolution.  That galaxies must sometimes cause multiple images
had long been expected \citep{1937ApJ....86..217Z}, and it had even
been argued that the phenomenon could help measure cosmological
parameters \citep{1964MNRAS.128..307R,1966MNRAS.132..101R}, but
apparently nobody was expecting that lenses would need detailed
modelling.  The first observations, however, immediately stimulated
models.  The reason for that lies in the image separation. Recall that
image separations are of order of the angular Einstein radius
\begin{equation}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2},
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  A lensing
galaxy with $M\sim10^{11}M_\odot$ at $10^9\rm\,pc$ would cause image
separations of $\sim1''$, which is comparable to the size of the
galaxy; typically the lensed images are seen through the galaxy halo.
Hence, the lensed images depend on the detailed mass distribution of
the lensing galaxy, and to uncover that dependence, galaxy lenses
require models.

Since those early discoveries, more than 400 secure lenses are now
known. Modelling of the mass distribution is part of any research
using lenses, but so far no modelling study has spanned all known
lenses.  The largest single one \citep{2009ApJ...703L..51K} models 58
separate lenses to infer the distribution of dark matter around
galaxies.  In other work, \cite{2011ApJ...740...97L} combined lens
models of 21 galaxies with models of their stellar populations, to
find the relation between stars and dark matter, and
\cite{2014MNRAS.437..600S} modelled 18 time-delay lenses together to
infer cosmological parameters.

@@

Consider a lens at some redshift $z_L$ and let $(x,y)$ be planar
coordinates at the lens, transverse to the line of sight.  Let
$\Sigma(x,y)$ be the mass distribution.  It is a mass per unit area,
i.e., density projected along the line of sight.  It is often given in
a dimensionless form
\begin{equation} \label{eq:kappa}
\kappa(x,y) \propto \Sigma(x,y)
\end{equation}
called the convergence.  Let there be light, in the form of a more
distant source, at redshift $z_S$, behind point $(x_s,y_s)$ on the
lens.

@@

After examining this feedback, the user can choose to save the
model to the \spl archive, at which point it is assigned an unique
URL.  He can also modify his input and try again, or discard the
attempt altogether.  After archiving, there can be discussion among
modellers, through the \sw forum or by any other means, and revision
of the model.  This is achieved simply by sharing the model's
URL; following its hyperlink takes one to the \spl app, pre-loaded
with the correct data image and input spaghetti. Any archived model
can be revised by any user: they can modify the spaghetti
configuration slightly or drastically, or change options like the size
of the mass tiles. Particularly interesting lens candidates lead to
trees of models in this way.  Discussion among modellers tends to
prune a model tree, focusing attention on the most interesting
models.\footnote{See ``Collaborative gravitational lens
modelling\dots'' in {\tt http://letters.zooniverse.org} for an example.}
% PJM: sadly this Letter is no longer available, Julianne took it
% down. We can post the talk URL instead.
% PS: Letter is available again now.
%\footnote{See the ``collaborative modelling'' thread at
% \url{ http://talk.spacewarps.org} for an example.}
